\SW: I. Crowd-sourcing the Discovery of Gravitational Lenses

### 2015-05-14 doc/sw-system.tex

Strong gravitational lensing -- the formation of multiple, magnified images of
background objects due to the deflection of light by  massive foreground
objects -- is a very powerful astrophysical tool, enabling a wide range of
science projects. The image separations and distortions provide information
about the mass distribution in the lens \citep[e.g.][]{AugerEtal2010,
SonnenfeldEtal2012,MoreEtal2012,SonnenfeldEtal2015}, including on sub-galactic scales
\citep[e.g.][]{Dalal+Kochanek2002,VegettiEtal2010,HezavehEtal2013}. Any strong
lens can provide magnification of a factor of 10 or more, providing a deeper,
higher resolution view of the distant universe through these ``cosmic
telescopes'' \citep[e.g.][]{StarkEtal2008,NewtonEtal2011}. Lensed quasars
enable cosmography via the time delays between the multiple images'
lightcurves \citep[e.g.][]{TewesEtal2013,SuyuEtal2013}, and study of the
accretion disk itself through the microlensing effect
\citep[e.g.][]{PoindexterEtal2008}. All of these investigations would
benefit from being able to draw from a larger and/or more diverse sample of lenses.

In the last decade the number of these rare cosmic alignments known
has increased by an order of magnitude, thanks to searches carried out in
wide field surveys, such as
CLASS \citep[e.g.]{BrowneEtal2003}, SDSS \citep[e.g.][]
{BoltonEtal2006,AugerEtal2010b,TreuEtal2011,InadaEtal2012,
HennawiEtal2008,BelokurovEtal2009,DiehlEtal2009,FurlanettoEtal2013},
\cfhtls \citep[e.g.][]{MoreEtal2012,GavazziEtal2014}, Herschel
\citep[][]{NegrelloEtal2014} and SPT \citep[e.g.][]{VieiraEtal2013}, among
others.  As the number of known lenses has increased, new types have been
discovered, leading to entirely new investigations. Compound lenses
\citep{GavazziEtal2008,CollettEtal2012} and lensed supernovae
\citep{QuimbyEtal2014,KellyEtal2015} are good examples of this.

Strong lenses are expensive to find, because they are rare. The highest purity
searches to date have made use of relatively clean signals such as the
presence of emission or absorption features at two distinct redshifts in the
same optical spectrum \citep[e.g.][]{BoltonEtal2004}, or the strong
``magnification bias'' towards detecting strongly-lensed sources in the sub-mm
waveband \citep[e.g.][]{NegrelloEtal2010}. Such searches have to yield pure samples,
because they require expensive high resolution imaging follow-up; consequently
they have so far produced yields of only tens to hundreds of lenses. An alternative
approach is to search images already of sufficiently high resolution and colour
contrast, and confirm the systems as gravitational lenses by modeling the
survey data themselves \citep[][]{MarshallEtal2009}. Several square degrees of
HST images have been searched, yielding several tens of galaxy-scale lenses
\citep[e.g.][]{MoustakasEtal2007,Jackson2008,MoreEtal2011,
PawaseEtal2014}. Similarly, searches of over a hundred square degrees of CFHT
Legacy Survey (\cfhtls) ground-based imaging, also with sub-arcsecond image quality,
have revealed a smaller number of wider image separation group-scale systems
\citep[e.g.][]{CabanacEtal2007,MoreEtal2012}. Detecting galaxy-scale lenses
from the ground is difficult, but feasible albeit with lower efficiency and requiring
HST or spectroscopic follow-up to confirm the candidates as lenses
\citep[e.g.][]{GavazziEtal2014}.

@@

A screenshot of the \SW classification interface (CI) is shown in
\Fref{fig:screenshot}. The CI is the centrepiece of the \SW website,
\texttt{http://spacewarps.org}; the web application is written in coffeescript,
css and html and follows the general design of others written by the Zooniverse
team.\footnote{The \SW web application code is open source and can be accessed
from \texttt{https://github.com/zooniverse/Lens-Zoo}.  The project was renamed
during development to avoid the question of  who ``Len'' is.} The focus of the
CI is a large display of a pre-prepared PNG image of the ``subject'' being
inspected. When the image is clicked on by the volunteer, a marker symbol
appears at that position.  Several markers can be placed, and they can be
removed by the classifier if they change their mind.

The next image moves rapidly in from a queue formed at the right hand side of
the screen when the ``Finished marking'' button is pressed. Gravitational lenses
are rare: typically, most of the images will not contain a lens candidate, and
these need to be quickly rejected by the inspector. The queue allows several
images to be pre-loaded while the volunteer is classifying the current subject,
and the rapid movement is designed to encourage volunteers to classify rapidly.
After each classification, the positions of the markers are written out to the
classification database, in an entry that also stores metadata including the ID
of the subject, the username of the volunteer  (or IP address if thy are not
logged in), and a timestamp.

For the more interesting subjects, the CI offers two features that enable their
further investigation. The first is the ``Quick Dashboard''
(QD), a more advanced image viewer. This allows the viewer to compare three
different contrast and colour balance settings, to help bring out subtle
features, and to pan and zoom in on interesting regions of the image to assess
small features. Markers can be placed in the Quick Dashboard just the same as
in the main CI image viewer. The second is a link to that subject's page in
the project discussion forum (known as
``\Talk\footnote{\texttt{http://talk.spacewarps.org}}'').
Here, volunteers can discuss the features they have seen either
before they submit their classification, or after, if they ``favourite'' the
subject. There is no ``back'' button: each volunteer may only classify a given
subject once. However, the presence of an option to see what others think
about any given subject before submitting your own classification means that
the classifications may not be strictly independent; however, the advantage of this
system is that volunteers can learn from others what constitutes a good lens
candidate. This is not, however, the primary educational resource; we describe
the explicit training that we provide for the volunteers in the next section.
The FITS images of the subject can be further explored with Zooniverse
``Dashboard Tools,'' which include a
more powerful image viewer that enables dynamic
variation of the colour balance and contrast (stretch), and for any given view
to be shared via unique URL back to \Talk. The FITS image viewing in both the
Quick Dashboard and the Dashboard Tools are enabled by the javascript library
\fitsjs \citep{fitsjs}.

@@

How fast does the \SW crowd classify subjects? Each software agent records the timestamp
of each classification its volunteer makes; by measuring the time lags between
successive classifications, we can make estimates of the crowd's classification
and contribution speed. We plot the former quantity for both \StageOne and
\StageTwo in \Fref{fig:results:speed}, normalizing the speed and time axes to
their respective totals. The fractional classification rates
in each stage of the survey fall off in approximately the same way, despite the
factor of nearly 20 difference in crowd size. \StageOne (consisting of
$\sim430,000$ subjects) was completed in around 5000 hours (by some 37,000
participants), with classification rates in the first few days averaging at
$\sim10^4$ per hour. This was stimulated by various forms of advertising (press
releases, and emails to registered Zooniverse users). The asymptotic
classification rate was  around ten times lower, at $\sim10^3$ per hour.
Interestingly, the average skill was found to be approximately constant over the
lifetime of \StageOne, leading to a contribution rate that tracks closely the
classification rate. (A 10\% increase in average skill was seen over the first
40 days, and a decrease of the same amount in the last 40 days -- not enough to
cause a significant difference between classification and contribution rate
behavior, but perhaps reflecting a learning period at the start and then a
decrease
in participation later on.) \StageTwo (3400 subjects) was completed in around 600
hours by around 2000 participants, with classification rates starting at
$~\sim2000$ per hour and decaying to $\sim100$ per hour.

@@

\begin{minipage}{\linewidth}
  \begin{minipage}[t]{0.47\linewidth}
    \begin{minipage}{0.46\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0000fgj_gri.pdf}
    \end{minipage}\hfill
    \begin{minipage}{0.50\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0000fgj_stage1_trajectory.pdf}
    \end{minipage}
  \end{minipage}\hfill
  \begin{minipage}[t]{0.47\linewidth}
    \begin{minipage}{0.46\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0003j2s_gri.pdf}
    \end{minipage}\hfill
    \begin{minipage}{0.50\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0003j2s_stage1_trajectory.pdf}
    \end{minipage}
  \end{minipage}
\end{minipage}
\caption{Illustrative examples of
false negatives (blue) and false positives (red) from the classification of
the \SW$-$\cfhtls training set. The trajectory plot for each of the 4 subjects is
shown to the right of its image. Lefthand panels show the most common types of
trajectories: random walks with short steps resulting from no high skill classifiers
viewing the subject. The righthand panels show examples where higher skill
classifiers have been involved, causing larger, more efficient ``kicks''
(top) but also catastrophic mis-classifications (bottom right).
Note that both these
subjects were {\it very nearly} detected but didn't quite reach the threshold
set (marked by the blue dashed line). Insets for the false
negatives show the lens feature. The inset for the lower righthand panel shows
where most volunteers believed a lens was located. There is no corresponding inset
for the lower left panel because no particular feature stood out to the
classifiers. Instead, the volunteers clicked many different features.}
\label{fig:discuss:performance:trajectories}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Some of  the mistakes made by reasonably high skill classifiers working at high
speed could have perhaps been corrected by those classifiers themselves, had
they had access to a ``go back'' option. While clearly enabling error reduction,
we might worry about such a mechanism enabling classifiers to have a guess, then
go and read the \Talk discussion, and then return to the classification
interface to change their mind to agree with the forum. In fact, this is already
possible on the current interface -- and indeed, it is not obvious that such
correlations would necessarily lead to poorer system performance. There may well
also be effects on volunteer confidence; it may be that encouraging this sort of
checking would result in increased and not necessarily productive caution. This
could be tested by presenting a fraction of the volunteers with a version of the
site that actively suggested that they take this approach, and then tracking the
relative performance of ``collaborative'' classifications compared with
independent ones. We leave the exploration of this to further work.

@@

However, identification failures (around 60\%, from inspection of a random sample)
seem to be due to the statistical noise associated with
the many short, semi-random direction kicks arising from classifications made by
low skill, inexperienced
classifiers (\Sref{sec:results:crowd}).
While improved training will help reduce this effect, it is here
that targeted task assignment could also help improve system performance, by
bringing higher skill volunteers in when needed. We
advertised \StageTwo of the current project to all registered users; it was
taken up by \StageOne veterans with a broad range of skill, and also picked up a
significant number of new users who did not have enough time to acquire high
skill (since \StageTwo was quite short): of the 1964 volunteers who took part in
the \StageTwo classification round, only 774 were veterans from \StageOne.

>>>>

Strong gravitational lensing -- the formation of multiple, magnified images of
background objects due to the deflection of light by  massive foreground
objects -- is a very powerful astrophysical tool, enabling a wide range of
science projects. The image separations and distortions provide information
about the mass distribution in the lens \citep[e.g.][]{AugerEtal2010,
SonnenfeldEtal2012,MoreEtal2012,SonnenfeldEtal2015}, including on sub-galactic scales
\citep[e.g.][]{Dalal+Kochanek2002,VegettiEtal2010,HezavehEtal2013}. Any strong
lens can provide magnification of a factor of 10 or more, providing a deeper,
higher resolution view of the distant universe through these ``cosmic
telescopes'' \citep[e.g.][]{StarkEtal2008,NewtonEtal2011}. Lensed quasars
enable cosmography via the time delays between the lightcurves of multiple images \citep[e.g.][]{TewesEtal2013,SuyuEtal2013}, and study of the
accretion disk itself through the microlensing effect
\citep[e.g.][]{PoindexterEtal2008}. All of these investigations would
benefit from being able to draw from a larger and/or more diverse sample of lenses.

In the last decade the number of these rare cosmic alignments known
has increased by an order of magnitude, thanks to searches carried out in
wide field surveys, such as
CLASS \citep[e.g.]{BrowneEtal2003}, SDSS \citep[e.g.][]
{BoltonEtal2006,AugerEtal2010b,TreuEtal2011,InadaEtal2012,
HennawiEtal2008,BelokurovEtal2009,DiehlEtal2009,FurlanettoEtal2013},
\cfhtls \citep[e.g.][]{MoreEtal2012,GavazziEtal2014}, Herschel
\citep[][e.g.]{NegrelloEtal2014} and SPT \citep[e.g.][]{VieiraEtal2013}, among
others.  As the number of known lenses has increased, new types have been
discovered, leading to entirely new investigations. Compound lenses
\citep{GavazziEtal2008,CollettEtal2012} and lensed supernovae
\citep{QuimbyEtal2014,KellyEtal2015} are good examples of this.

Strong lenses are expensive to find, because they are rare. The highest purity
searches to date have made use of relatively clean signals such as the
presence of emission or absorption features at two distinct redshifts in the
same optical spectrum \citep[e.g.][]{BoltonEtal2004}, or the strong
``magnification bias'' towards detecting strongly-lensed sources in the sub-mm/mm
waveband \citep[e.g.][]{NegrelloEtal2010}. Such searches have to yield pure samples,
because they require expensive high resolution imaging follow-up; consequently
they have so far produced yields of only tens to hundreds of lenses. An alternative
approach is to search images already of sufficiently high resolution and colour
contrast, and confirm the systems as gravitational lenses by modeling the
survey data themselves \citep[][]{MarshallEtal2009}. Several square degrees of
HST images have been searched, yielding several tens of galaxy-scale lenses
\citep[e.g.][]{MoustakasEtal2007,Jackson2008,MoreEtal2011,
PawaseEtal2014}. Similarly, searches of over a hundred square degrees of CFHT
Legacy Survey (\cfhtls) ground-based imaging, also with sub-arcsecond image quality,
have revealed a smaller number of wider image separation group-scale systems
\citep[e.g.][]{CabanacEtal2007,MoreEtal2012}. Detecting galaxy-scale lenses
from the ground is difficult, but feasible albeit with lower efficiency and requiring
HST or spectroscopic follow-up to confirm the candidates as lenses
\citep[e.g.][]{GavazziEtal2014}.

@@

A screenshot of the \SW classification interface (CI) is shown in
\Fref{fig:screenshot}. The CI is the centrepiece of the \SW website,
\texttt{http://spacewarps.org}; the web application is written in coffeescript,
css and html and follows the general design of others written by the Zooniverse
team.\footnote{The \SW web application code is open source and can be accessed
from \texttt{https://github.com/zooniverse/Lens-Zoo}.  The project was renamed
during development to avoid the question of  who ``Len'' is.} The focus of the
CI is a large display of a pre-prepared PNG image of the ``subject'' being
inspected. When the image is clicked on by the volunteer, a marker symbol
appears at that position.  Multiple markers can be placed, and they can be moved or
removed by the classifier if they change their mind.

The next image moves rapidly in from a queue formed at the right hand side of
the screen when the ``Next'' or ``Finished Marking'' button is pressed (If a marker is placed somewhere in the image, the Next button changes to a ``Finished marking'' button). Gravitational lenses
are rare: typically, most of the images will not contain a lens candidate, and
these need to be quickly rejected by the inspector. The queue allows several
images to be pre-loaded while the volunteer is classifying the current subject,
and the rapid movement is deliberately designed to encourage volunteers to classify quickly. There is no ``back'' button in the CI: each volunteer may only classify a given
subject once and subjects cannot be returned to after the `Next or Finished marking' button has been pressed (this can be a source of missed lens candidates when classifying at speed). Note that by `classify' we mean that we interpret no markers being placed as a rejection, and placing at least one marker to mean a possible lensing event has been identified.
After each classification, the positions of any markers are written out to the
classification database, in an entry that also stores metadata including the ID
of the subject, the username of the volunteer (or IP address if thy are not
logged in), and a timestamp.

For the more interesting subjects, the CI offers two features that enable their
further investigation. The first is the ``Quick Dashboard''
(QD), a more advanced image viewer. This allows the viewer to compare three
different contrast and colour balance settings, to help bring out subtle
features, and to pan and zoom in on interesting regions of the image to assess
small features. Markers can be placed in the Quick Dashboard just the same as
in the main CI image viewer. The second is a link to that subject's page in
the project discussion forum (known as
``\Talk\footnote{\texttt{http://talk.spacewarps.org}}'').
Here, volunteers can discuss the features they have seen either
before they submit their classification, or after, if they ``favourite'' the
subject. The option of reading other opinions on any given image or lens candidate before submitting a classification means that
the classifications may not be strictly independent; however, the advantage of this
system is that volunteers can learn from others what constitutes a good lens
candidate (we are not able to track who visited talk before pressing ``Finished marking''). This is not, however, the primary educational resource; we describe
the explicit training that we provide for the volunteers in the next section.
The FITS images of the subject can be further explored with Zooniverse
``Dashboard Tools,'' which include a
more powerful image viewer that enables dynamic
variation of the colour balance and contrast (stretch), and for any given view
to be shared via unique URL back to \Talk. The FITS image viewing in both the
Quick Dashboard and the Dashboard Tools are enabled by the javascript library
\fitsjs \citep{fitsjs}.

@@

How fast does the \SW crowd classify subjects? Each software agent records the timestamp
of each classification its volunteer makes; by measuring the time lags between
successive classifications, we can make estimates of the crowd's classification
and contribution speed. We plot the former quantity for both \StageOne and
\StageTwo in \Fref{fig:results:speed}, normalizing the speed and time axes to
their respective totals. The fractional classification rates
in each stage of the survey fall off in approximately the same way, despite the
factor of nearly 20 difference in crowd size. \StageOne (consisting of
$\sim430,000$ subjects) was completed in around 5000 hours (by some 37,000
participants), with classification rates in the first few days averaging at
$\sim10^4$ per hour. This was stimulated by various forms of advertising (press
releases, and emails to registered Zooniverse users). The asymptotic
classification rate was  around ten times lower, at $\sim10^3$ per hour.
Interestingly, the average skill was found to be approximately constant over the
lifetime of \StageOne, leading to a contribution rate that tracks closely the
classification rate. (A 10\% increase in average skill was seen over the first
40 days, and a decrease of the same amount in the last 40 days -- not enough to
cause a significant difference between classification and contribution rate
behaviour, but perhaps reflecting a learning period at the start and then a
decrease
in participation later on.) \StageTwo (3400 subjects) was completed in around 600
hours by around 2000 participants, with classification rates starting at
$~\sim2000$ per hour and decaying to $\sim100$ per hour.

@@

\begin{minipage}{\linewidth}
  \begin{minipage}[t]{0.47\linewidth}
    \begin{minipage}{0.46\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0000fgj_gri.pdf}
    \end{minipage}\hfill
    \begin{minipage}{0.50\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0000fgj_stage1_trajectory.pdf}
    \end{minipage}
  \end{minipage}\hfill
  \begin{minipage}[t]{0.47\linewidth}
    \begin{minipage}{0.46\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0003j2s_gri.pdf}
    \end{minipage}\hfill
    \begin{minipage}{0.50\linewidth}
      \centering\includegraphics[width=\linewidth]{ASW0003j2s_stage1_trajectory.pdf}
    \end{minipage}
  \end{minipage}
\end{minipage}
\caption{Illustrative examples of
false negatives (blue) and false positives (red) from the classification of
the \SW$-$\cfhtls training set. The trajectory plot for each of the 4 subjects is
shown to the right of its image. Lefthand panels show the most common types of
trajectories: random walks with short steps resulting from no high skill classifiers
viewing the subject. The righthand panels show examples where higher skill
classifiers have been involved, causing larger, more efficient ``kicks''
(top) but also catastrophic mis-classifications (bottom right).
Note that both these
subjects were {\it very nearly} detected but didn't quite reach the threshold
set (marked by the blue dashed line). Insets for the false
negatives show the lens feature. The inset for the lower righthand panel shows
where most volunteers believed a lens was located. There is no corresponding inset
for the lower left panel because no particular feature stood out to the
citizens. Instead, the volunteers clicked many different features.}
\label{fig:discuss:performance:trajectories}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Some of  the mistakes made by reasonably high skill classifiers working at high
speed could have perhaps been corrected by those classifiers themselves, had
they had access to a ``go back'' option. While clearly enabling error reduction, particularly those caused by simply skimming too fast,
we might worry about such a mechanism enabling classifiers to have a guess, then
go and read the \Talk discussion, and then return to the classification
interface to change their mind to agree with the forum. In fact, this is already
possible on the current interface -- and indeed, it is not obvious that such
correlations would necessarily lead to poorer system performance.
\flag{AV}{REPETITION OF
TALK FACTOR FROM EARLIER SECTION, IT"S NOT NEEDED TWICE}. There may well
also be effects on volunteer confidence; it may be that encouraging this sort of
checking would result in increased and not necessarily productive caution. This
could be tested by presenting a fraction of the volunteers with a version of the
site that actively suggested that they take this approach, and then tracking the
relative performance of ``collaborative'' classifications compared with
independent ones. We leave the exploration of this to further work.

@@

However, identification failures (DEFINE WHAT YOU MEAN BY IDENTIFICATION FAILURES around 60\%, from inspection of a random sample)
seem to be due to the statistical noise associated with
the many short, semi-random direction kicks arising from classifications made by
low skill, inexperienced
classifiers (\Sref{sec:results:crowd}). ARE YOU SAYING THAT THERE ARE MORE OF THESE THAN THE RIGHT COLUMN EXAMPLES?  WE SHOULD GIVE NUMBERS.
While improved training will help reduce this effect, it is here
that targeted task assignment could also help improve system performance, by
bringing higher skill volunteers in when needed. We
advertised \StageTwo of the current project to all registered users; it was
taken up by \StageOne veterans with a broad range of skill, and also picked up a
significant number of new users who did not have enough time to acquire high
skill (since \StageTwo was quite short): of the 1964 volunteers who took part in
the \StageTwo classification round, only 774 were veterans from \StageOne.
